```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Identifying Schools with Biggest Differences in Math Test Scores

### Introduction

In this analysis, we will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. We'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores.

This analysis can be useful for journalists to understand which schools are performing better or worse than expected, based on the socioeconomic status of their student population. By identifying these schools, targeted interventions and support can be provided to help improve student outcomes - and to provide some context on comments by policymakers.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```
### Task: Create a Scatterplot and Describe the Results (2 points)

Next, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```

The scatterplot shows the overall trend of the relationship between the FARMS percentage and the math test scores. The red line represents the linear regression line, which helps us understand the average relationship between the two variables. Based on what you see here, what's the relationship between FARMS percentage and math test scores?

There doesn't seem to be much of a clear relationship here, though there are outliers that show a positive relationship between FARMS percentage and math test scores. It's fascinating how spread out the scores are outside of a few select dots that clearly show the correlation. A deeper analysis seems to be necessary to explore deeper, though initially I simply don't see a clear correlation from this chart. 


### Task: Linear Regression Model (3 points)

Next, we'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

There are three things we need to examine here: the p-value, the Adjusted R-squared value and the coefficients. You know what a p-value is and how to evaluate it. The Adjusted R-squared is a measure of how much of the difference between schools' math scores can be explained by their FARMS percentage. You don't need to multiply that figure by anything: that's a percentage. The closer the adjusted R-squared is to 100, the more of the outcome the independent variable explains. If both the p-value and adjusted R-squared values are significant, we can look to the coefficients.

In the middle, you can see the (Intercept) is 59.57897 and the FARMS coefficient is 0.12349. Remember high school algebra? Remember learning the equation of a line? Remember swearing that learning y=mx+b is stupid because you’ll never need it again? Surprise. It’s useful again. In this case, we could try to predict a school's score math score by using y=mx+b. In this case, y is the math score, m is 0.12349 and b is 59.57897. So we would multiply a school's FARMS percentage by 0.12349 and then add 59.57897 to it. The result would tell you what the math score would be, according to our model.

Based on what you've learned so far, is this model worth using? What does it tell us about the nature of the relationship between FARMS percentage and math scores?

This model does seem worth using. If nothing else it at least shows statistical significance in multiple ways — through things like the p-value, R-squared value, residuals and such. In this case it doesn't seem to show much of a statistical significance in the relationship between FARMS percentage and math scores — even the first thing I looked at (the p-value) corroborates the idea that there is not a statistical significance here. 

### Task: Identifying Schools with Biggest Residuals (2 points)

To find the schools with the biggest differences between their actual and expected math test scores, we'll extract the residuals from the linear regression model. The residuals represent the difference between the actual and predicted values. Residuals can tell you several things, but most important is if a linear model the right model for your data. If the residuals appear to be random, then a linear model is appropriate. If they have a pattern, it means something else is going on in your data and a linear model isn’t appropriate.

Residuals can also tell you who is under-performing and over-performing the model. And the more robust the model – the better your r-squared value is – the more meaningful that label of under or over-performing is. 

We'll then use the arrange() function to sort the data frame by the absolute value of the residuals.

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```
The schools with the biggest positive residuals are performing better than expected, while the schools with the biggest negative residuals are performing worse than expected, based on their FARMS percentage. Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores?

There are some large differences in predicted results and actual math scores. For example, Whitemarsh elementary school had 26.1 percent of students at least meet math expectations, but was predicted to have nearly 61 percent in that metric. The residual was nearly 35, the highest total in this chart. Gulph elementary school follows with a residual just over 30. There is one other school with a difference of over 30. 

## Task: Evaluate the model (3 points)

A school board member has said at a meeting that elementary schools with fewer students on free and reduced price meals do better on standardized math tests. Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

From the model I would draw that there isn't significant evidence to show that having fewer students on free and reduced price meals leads to better scores on standardized math tests. With the model, and all the data taken from the linear regression model — like p-value, R-squared value, etc — it demonstrates a lack of statistical significance between the meals and math scores. Frankly, the data would make me more hesitant to pitch a story about the correlation since there just isn't a clear significance. Cherrypicking data isn't a smart and ethical move, and that seems to be the case in the school board member's claim at the meeting. 
