---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}

set.seed(42) # Setting seed for reproducibility
population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

# View the first few rows
head(population)

```



### Task 3: Sample 100 rows of data from the dataset

```{r}

sample_100 <- population |> 
  sample_n(100)

sample_50 <- population |> 
  sample_n(50)

sample_10 <- population |> 
  sample_n(10)


```


### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample_100 |> summarize(mean = mean(age), sd = sd(age))
sample_50 |> summarize(mean = mean(age), sd = sd(age))
sample_10 |> summarize(mean = mean(age), sd = sd(age))
```



### Task 5: Calculate the mean age and standard deviation of the population

```{r}
population |> summarize(mean = mean(age), sd = sd(age))
```

### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The means and standard deviations are very similar, with minimal difference between the sample and population staitstics. Even when I change the sample to 50, it is still pretty similar. However, once I narrow down the sample to 10, it drastically changes. The mean goes up by nearly 10 while the standard deviaition shifts by about 2 — in the context of this dataframe, that is a massive shift. A sample of 50 still represents the overall dataset well, but once it drops to 10, the representation is clearly skewed far from the true mean and standard deviaition. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
