---
title: "Prince George's County Car Theft Spike Analysis"
author: "Jake Kauderer"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Introduction

This analysis examines the impact of a significant spike in car theft rates in 2023 in Prince George's County, and whether that spike has led to a sustained higher rate of car theft in the county. We'll use Analysis of Variance (ANOVA) to determine if theft rates differ significantly between pre-spike (before 2023), spike (2023), and post-spike (2024-present) periods.

## Data Preparation and Exploration

### Loading Libraries and Data

```{r load-libraries}
# Load required libraries
library(tidyverse)
library(lubridate)
```

```{r load-data}
# Import data
theft_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the first few rows
head(theft_data)

# Get a summary of the data
summary(theft_data)
```


### Get the data how we want it and calculate the rates

**Task 1** Replace REPLACE_ME with the variables you need to calculate the theft_rate (2 points)

```{r process-data}
# Calculate theft rate and create time period variable
theft_data <- theft_data |>
  mutate(
    date = as.Date(month),
    year = year(date),
    month_number = month(date),
    theft_rate = (total / population) * 100000,
    time_period = case_when(
      year < 2023 ~ "Pre-spike",
      year == 2023 ~ "Spike year",
      year > 2023 ~ "Post-spike"
    )
  )

# Create summary by time period
period_summary <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_thefts = mean(total),
    mean_rate = mean(theft_rate),
    median_rate = median(theft_rate),
    sd_rate = sd(theft_rate),
    n = n(),
    .groups = "drop"
  )

# Display the summary
knitr::kable(period_summary, digits = 2,
             caption = "Summary of Car Theft Rates by Time Period")
```

**Task 2** Describe the results of calculating the rates for each period. What do they suggest? (3 points)

The results show that the spike year, as expected, had by far the highest theft rate. While the theft rate went down post-spike, the numbers were still well inflated from pre-spike numbers. This shows that the spike had lasting affects for years after, even if the rates decreased from the actual spike year.

## Visualizing the Spike Pattern

### Time Series Plot

```{r time-series}
# Time series plot showing the spike pattern
ggplot(theft_data, aes(x = date, y = theft_rate)) +
  geom_line() +
  geom_smooth(method = "loess", span = 0.2) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "red") +
  geom_vline(xintercept = as.Date("2024-01-01"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("2023-07-01"), y = max(theft_data$theft_rate) * 0.9, 
           label = "Spike Year", color = "red") +
  labs(title = "Monthly Car Theft Rates (2017-2025) with Spike Highlighted",
       x = "Date", y = "Thefts per 100,000 Population") +
  theme_minimal()
```

**Task 3** Describe the output of this chart and then give it a better, more active title that provides the lede (3 points)

The output of this chart demonstrates how massive the spike was. While the theft rate was already going up, the spike was a massive, massive increase, though it did eventually go down. The reader can see that the rate continued to go down for the majority of the following year, though it was still well above the prior year's rates. This chart corroborates the data viewed earlier, showing the huge implication of the spike — both in the year it occurred and even for the year after. With that said, it does seem like the rate is going back down, though to what extent is to be seen.

Title: A Spike in Car Theft Rates in Prince George's County has had Lasting Effects
Lede: Car Theft rates spiked in Prince George's County in 2023. Despite rates mainly falling since, they remain at a higher clip than the years before the spike.

Now let's take a look at average theft rates.

### Yearly Average Theft Rates

```{r yearly-means}
# Calculate yearly means
yearly_means <- theft_data |>
  group_by(year) |>
  summarize(
    mean_rate = mean(theft_rate),
    .groups = "drop"
  )

# Plot yearly means
ggplot(yearly_means, aes(x = year, y = mean_rate)) +
  geom_line(size = 1) +
  geom_point(size = 3, color = "blue") +
  geom_point(data = yearly_means |> filter(year == 2023), 
             aes(x = year, y = mean_rate), size = 4, color = "red") +
  labs(title = "Average Annual Car Theft Rates",
       x = "Year", y = "Average Thefts per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = 2017:2025)
```

## One-way ANOVA Analysis

### Hypothesis Formulation

For our ANOVA test, we'll compare the mean theft rates between the three time periods. First we need to create our hypotheses.

**Task 4** Fill in the hypotheses (2 points)

- **Null Hypothesis (H0)**: Theft rates do not significantly differ in the respective periods — pre-spike, spike and after-spike. 

- **Alternative Hypothesis (H1)**: Theft rates significantly differ in the respective periods — pre-spike, spike and after-spike.

### One-way ANOVA Test

```{r anova-test}
# Perform one-way ANOVA on theft rates by time period
period_anova <- aov(theft_rate ~ time_period, data = theft_data)
summary(period_anova)
```

**Task 5** Describe the result: are there real differences in the average rates between the time periods? Is this result significant? (5 points)

There are real differences in the average rates between the time periods. The F-value is large (218), which demonstrates that the differences between time periods outweighs the variation within each period. The p-value is low, indicating the differences between rates over respective time periods are statistically significant. The three asterisks indicate a significance at the 0.001 level, helping to show that there are real differences in the average rates between the time periods.

## Compare the differences between specific periods

Run our post-anova Tukey test and examine the results. 

```{r posthoc}
# Tukey's HSD test
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```


```{r period-means}
# Visualize time period means with confidence intervals
period_rates <- theft_data |>
  group_by(time_period) |>
  summarize(
    mean_rate = mean(theft_rate),
    se = sd(theft_rate) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Ensure time period is ordered correctly
period_rates$time_period <- factor(period_rates$time_period, 
                                   levels = c("Pre-spike", "Spike year", "Post-spike"))

ggplot(period_rates, aes(x = time_period, y = mean_rate, fill = time_period)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  labs(title = "Car Theft Rates by Time Period with 95% Confidence Intervals",
       x = "", y = "Mean Theft Rate per 100,000") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This analysis has examined whether there are significant differences in car theft rates between the pre-spike period (before 2023), the spike year (2023), and the post-spike period (after 2023).

**Task 6** Describe the results of the Tukey's test and chart showing confidence intervals. Is there a significant difference between the three periods? In particular, is the difference between the pre-spike and post-spike periods real and significant? Write a headline and lede reporting these findings, focusing on the pre- and post-spike periods. (5 points)

There is a significant difference between the three periods, with the pre-spike being by far the lowest rate and then the spike being the highest, followed by the post-spike in the middle. The difference between the pre-spike and post-spike periods is real and significant. While rates went down after the spike, the post-spike period is still seeing much higher rates than the years before the spike.

Headline: A 2023 spike in car thefts in Prince George's county has had lasting effects
Lede: The mean rate of car thefts pre-2023 was about 200 less compared to post-spike numbers, despite a decrease in thefts since the spike.
