---
title: "Diversity Index for Maryland Cities"
author: "Jake Kauderer"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?

This data contains many metrics about diversity in cities, like population by race and the amount of two-plus races in cities. Baltimore City stands out as diverse, it has a far higher quantity of "two-plus races" than any other city. Germantown and Columbia have lower quantities than Baltimore City, but they still stand out as extremely diverse. 

```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?

Summerfield has the highest race probability. This makes complete sense, as it only has 126 "two-plus races" — that number is the second lowest of any city. The low diversity correlates with the higher race probability sum.

```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```


## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?

I notice that the cities with the lowest Hisp_Prob scores actually have higher percentage of Hispanics. Like the cities with the lowest three Hisp_Prob scores all have at least a 40 percent Hispanic population — only one other city has 40 percent or above. This likely holds true because Hisp_Prob isn't just measuring how many people are Hispanic, it's measuring the divide in the population between Hispanics and Non-Hispanics.

```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?

The top 3 most diverse cities according to the index are Langley Park, Wheaton and Adelphi. Adelphi and Langley Park are both in Prince Groge's County, while Wheaton is in Montgomery County. The least diverse city is Lake Shore which is in Anne Arundel County.  

```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

Headline: Hispanics in Maryland live in extremely diverse cities 
Lede: Langley Park is Maryland’s most Hispanic community — about 81% of its population is Hispanic — and it’s also the most diverse. With a Diversity Index of 97, there’s a 97% chance that two randomly chosen residents will be from different racial or ethnic backgrounds.

Details: I'd talk to residents in Langley Park and ask them about the diversity that they notice — getting anecdotes about how they feel could add to the story. It would be fascinating to talk to Hispanics in the area — finding out how they feel about their large presence being there, yet it still being such a diverse community. I would also want to talk to teachers/students at schools and ask how the diversity plays an impact in academia. It would also be beneficial to talk with government workers, like cops, the mayor, etc. I think gauging their views on the huge diversity in the city would be smart since many people would likely be intrigued by how they feel. I'd also ask all of these people the question: has diversity always been this high? If not, why is it now? Finding this out could add a layer to the story, especially if it hasn't always been such a diverse area.
