---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Jake Kauderer"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

We'll do what we did in class, but with a twist: instead of Montgomery County, Pa., you'll be working with data from Montgomery County, Md. Same basic idea, though: linear regression looking at free and reduced-price meals and math test scores.

Just as in the exercise, you will create a scatterplot to visualize the relationship between the FARMS (Free and Reduced-price Meal System) percentage and the percentage of students who met or exceeded math expectations. You'll then use linear regression to identify the schools that show the biggest differences between their actual and expected math test scores and evaluate the results of your model.

### Loading the Data

First, let's load the data into a data frame using the read_csv() function from the tidyverse library.

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```
### Task: Create a Scatterplot and Describe the Results (5 points)

First, let's create a scatterplot to visualize the relationship between the FARMS percentage and the percentage of students who met or exceeded math expectations. We'll use ggplot2 to create the plot and add a linear regression line.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations", title = "Scatterplot of Math Expectations vs FARMS Percentage")
```
Based on what you see here, what's the relationship between FARMS percentage and math test scores? Compare this scatterplot to the one you made in the class activity.

Based on this scatterplot, the relationship between FARMS percentage and math test scores is negatively correlated. As the FARMS percentage increased, the percentage of students who met or exceeded math expectations decreased (mainly). The one we looked at in class was way more sporadic, with a few outliers that, if anything, showed a positive correlation. This scatterplot tells a completely different story with the clear negative correlation.


### Task: Linear Regression Model (7 points)

Next, you'll perform a linear regression analysis using the lm() function. We'll set the percentage of students who met or exceeded math expectations as the dependent variable, and the FARMS percentage as the independent variable. In other words, if we know the FARMS percentage for a school, how reliably can we predict its math score?

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)

summary(model)
```

Provide the model's output for the three statistics mentioned in the class activity.

P-value = 0.00000000000000022. R-squared value = 0.6734. Residuals =  Min: -28.0844, Max: 29.9939.

Based on what you've learned so far, what does it tell us about the nature of the relationship between FARMS percentage and math scores? Is this model worth exploring further? Why or why not?

This model is definitely worth further exploring as it shows there is a significance in the relationship between FARMS percentage and math scores — a contrast from the in-class assignment. Even just with a preliminary look at the p-value, a person can see that is is extremely low. That means there is some significance present. The R-squared value is higher than in the in-class assignment (closer to 100 means it is more significant in this context). The minimum and maximum residuals are pretty similiar to the activity we did in class, so further exploration is needed to gauge their significance. Overall, there definitely seems to be potential for a story here and further exploration would be smart. 

### Task: Identifying Schools with Biggest Residuals (3 points)

Let's calculate and examine the residuals next. 

```{r}
schools$predicted <- predict(model)

# Calculate the residuals
schools <- schools |> 
  mutate(residual = `% Met or Exceeded Math Expectations` - predicted)

# Sort the data frame by the absolute value of the residuals in descending order
schools <- schools |>
  mutate(abs_residual = abs(residual)) |>
  arrange(desc(abs_residual))

```

Look at the absolute values of the residuals; what are the largest differences between the predicted result and the actual math scores? Pick two schools with the largest residuals. Find the zip codes where they are located and using the CensusReporter.org website, describe the most interesting demographic data from each zip code.

The two schools with the largest residual differences of about 29 and 28, respectively, are Fox Chapel Elementary and Broad Acres Elementary. For Fox Chapel Elementary, one of the most interesting demographic pieces of information is that there's a pretty even distribution of White, Black, Asian and Hispanic people in the population. No real distinctions to make there, though the area does have a high amount of foreign-born residents (over double the rate of Maryland as a whole). Broad Acres Elementary has 53 percent Hispanic and 26 percent Black people. There is a much lower amount of Asian and White people in the population there.

## Task: Evaluate the model (5 points)

Describe what conclusions you'd draw from this model's results and how it would influence a story pitch about the school board member and math scores.

The model (especially the p-value and R-squared value) corroborate my initial thoughts from the scatterplot — there is a negative correlation between FARMS percentage and math test scores. The residual information also backs this point up — even just taking a quick glance a person can see that the schools with the highest FARMS percentages have a worse percentage of students who met or exceeded math expectations. Then it's the same vice-versa, helping corroborate the negative correlation between FARMS percentage and math test scores. A story pitch could be about that, as it's something that many students, teachers and (especially) parents would likely have great interest in. This assignment gives a much clearer route for a story, compared to the in-class one where there was minimal correlation and not much significance demonstrated from the model. 
