---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Jake Kauderer"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

Lower unemployment seems to correlate with a higher voter turnout. Also, while a bit inconsistent, it seems that generally the older groups are voting more (like most of the top turnout places have a median age in the 40s to 50s, while the bottom turnout places are more so in the 30s). It also seems like generally the more educated people vote more, though this is definitely a bit sporadic. There doesn't seem to be much correlation at all with income and turnout. In the scatterplots, unemployment, percent white and median age seem to have a relatively linear relationship with the dependent variable. Income, education and population seem to have a less clearly linear relationship.

```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

The model seems to predict voter turnout pretty well, with the residuals very close to 0. This signifies that the variables chosen are the right ones. Income seems to be relatively insignificant, there may be a slight relationship with voter turnout but it's pretty sporadic. Unemployment seems similar in terms of how sporadic it is. This is interesting because from the initial matrix I thought unemployment had more of a correlation with voter turnout. Overall, the model does seem useful as it helps further prove or unprove correlations. Also, the low residuals and low median signify that the variables chosen do work pretty well in the model.

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

For mine model 2 was by far the worst, this is interesting as I used income and college education for it. Would definitely want to follow up with this and find out exactly why this is. The R-squared value is around 12 which is extremely low. The 3rd and 4th model were equivalent in their usefulness, with a respectable R-squared value of about 45. These are more in the middle range, so some more research/experimentation with variables could be useful to potentially boost the R-squared value. 

```{r}
# Model 2
model2 <- lm(turnout ~ income + pct_bachelors_higher, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ median_age + pct_white - income, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ median_age + pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

This method could be utilized through its ability to analyze how different factors play a role in voter participation. Does lower unemployment in a town result in a higher voter turnout? How does race affect who votes? Such things aren't easy to gauge, but can be done in at least a somewhat effective way using this method. The ability to distinguish between factors is important, and this method helps with that. It also feels smart and relevant to find these things out since a lot of people would likely care about various factors that affect voter turnout, considering the unmatched importance of elections today. 
