---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

There are massive fluctuations in the values of salaries for white house workers. For example, a special policy advisor (Peter A. Devine) literally earns nothing, while a deputy assistant to the president earns 155,000. These huge deviations in salaries could cause issues in analysis, as there is just such a large range to go through. The one's that are particularly low, and even zero, stand out to me. It would be interesting to investigate further as to how that can be, especially when there are people in other positions making such high salaries.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(salary/564)
wh_salaries |> summarize(median = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
mean_salary <- wh_salaries |> summarize(mean = mean(salary))
median_salary <- wh_salaries |> summarize(median = median(salary))
min_salary <- wh_salaries |> reframe(min = min(salary))
max_salary <- wh_salaries |> reframe (min=max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary is $107,036.6, which seems pretty high -- honeslty I was expecting it to be lower considering there are some people who aren't paid at all. The median is 110,500 which seems about what you'd expect from this dataset, nothing too crazy to take from that. The minimum salary is $0, which is very interesting. It would be fascinating to talk to the people who make nothing and ask them how they deal with that. The max salary is $251,258 which is a lot of money, I'd love to talk to the person who makes that much. This data tells me there is a huge disparity in salary among workers at the white house. Both the minimum and maximum are clear outliers as none are close to the mean salary (over $100,000 difference). There are certainly outliers sprinkled throughout the dataset.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

The story I would go for is one about the huge disparities in workers at the white house. The next step would be talking to people from the white house, and certainly those who make $0 all the way to those who make over $200,000. I'd want to answer the question, why is there such a massive disparity in salaries among white house workers?
