---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 10000, property_rate_2019 = property2019/pop2019 * 10000, violent_rate_2020 = violent2020/pop2020 * 10000, property_rate_2020 = property2020/pop2020 * 10000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Baltimore had the highest rate of violent crime by a wide margin in 2019. That number decreased for 2020, though Baltimore still had the second highes rate (Ocean City's skyrocketed from 2019 to 2020). In terms of property crimes, Ocean City had the highest rate by far in 2019 -- that rate increased even more in 2020. The story here is about Ocean City, which clearly underwent some huge struggles, especially with such a drastic increase in both violent and property crime.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |>  mutate(violent = (violent_rate_2020 - violent_rate_2019 * 100), property = (property_rate_2020 - property_rate_2019 * 100))

```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The percentage change results are very mixed, some cities rates went up drastically while others went down. There were also a bunch that didn't change at all. I believe the outliers are the biggest story here -- think of Deer Park's huge reduction in property crime and Ocean City's huge increase in violent crime from 2019-2020. I believe it would be more fascinating to find out what the factors were in these massive changes, rather than just discussing the general trends in the rates during the span. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Ocean City's crime rates drastically increased from 2019-2020.

Ocean City was well behind Baltimore in terms of violent crime in 2019. 2020 changed that completely, with Ocean City's rate increasing by over 100 while Baltimore's rate went down. The rate of property crime in Ocean City also saw a major increase, with 120 more property crimes committed. Whether it be policy or other issues, Ocean City clearly had a rough stretch with multiple types of crimes. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

## Task 10: HW 5

Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?

```{r}
md_cities_percentile<-md_cities_rates |> mutate(percentile = ntile(violent_rate_2020, 100))
```
11 cities are in the 90th percentile, proven by the md_cities_percentile chart.

Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.

```{r}
md_cities_change |> summarize(violent_2020_IQR = IQR(violent_rate_2020))
```
These results do not change my story idea from the last assignment, as I believe Ocean City is still the most intriguing story to try and tell even with the new information on the 2020 violent crime rate. 


