---
title: "Final Project"
Jake Kauderer
output: html_document
---

Final Project: The Tests You Could Run

- To start, z-scores seem like they could be especially useful in analyzing this data (which on the surface doesn't seem to tell a lot since these stats are nearly identical in wins or losses). One possibility is to try and find outlier games (what a z-score could help tell) and analyze them. For example, there are some games where a team scores few goals (3 or below) or a lot (18 or above). I could look at these and try and find out what there's a high correlation in (are teams turning it over less, are they getting a ton of saves, are they shooting incredibly few or many shots, etc). My thought process is that on average there doesn't seem to be enough disparity to tell an insightful story, but possibly looking at those outliers could give something more. Another test I could possibly use is the chi-square test. This could tie back into my original idea for this project of testing if there is a high correlation between teams playing at home and winning (something I have not tested yet). It's a bit of a different idea from all the stats I used below, but it could be interesting, as I could use the chi-square test to find if there is a correlation or not. Another possibility is using ANOVA to find p-values — I could utilize this to maybe combine my ideas in a way? For example, what if I tracked if teams have a higher total of shots at home compared to the road? How about saves and turnovers? Not sure how this would exactly work, but it's an idea I had that could tell something interesting. These three tests, all of which we covered pretty recently, seem like they could be useful in helping me find a story to tell out of this lacrosse data. However, obviously you have more experience and I'm definitely interested to hear what you're thinking about where I'm at.  



Link to data (via Derek (big thanks)): https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mlax_matches.csv 

So this data is on lacrosse stats over the past five years into this week. The columns I chose to analyze were saves, ground balls, turnovers and shots to see if there's a heavy correlation in them, respectively, to wins. The results were very telling in that there was very little difference in them for wins versus losses. It's super interesting and makes me wonder what stat might tell more of a story. My question to you would be could I get metrics on different things like shot efficiency, faceoffs and more that might have a larger disparity? Faceoffs particularly interests me, if you could provide any insight on if I could get that data for the past five seasons/how I would go about it, that would be great. Also, do you think there's a story to tell here with the data being so similar for wins and losses with the four categories I analyzed? To me, the differences are so small that I'm not sure there would be, but maybe going about it from the perspective that things truly even out in the end could work? Not positive, would be interested to hear what you think. Also if you have any general recommendations for how I'm going about this, definitely feel free to share!


```{r} 
library(tidyverse)
```


Full transparency, chatGPT helped with code, on the syllabus it says it's permitted as long as it's disclosed and the prompt is given for respective codes.  

### Task 2: Load the data

```{r}
lacrosse_games2025 <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mlax_matches.csv")
```

For this, gave the dataset and then said "I have this dataset above with stats about mens lacrosse from the past five years. I want to gauge if a higher amount of saves actually correlates to wins, can you help me with what code to add to do that?"

```{r}
lacrosse_games2025 <- lacrosse_games2025 %>%
  mutate(win = if_else(result == "W", 1, 0))
```
```{r}
lacrosse_games2025 %>%
  group_by(result) %>%
  summarize(
    avg_saves = mean(saves, na.rm = TRUE),
    median_saves = median(saves, na.rm = TRUE),
    count = n()
  )
```
```{r}
ggplot(lacrosse_games2025, aes(x = saves, fill = result)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  labs(
    title = "Distribution of Saves in Wins vs. Losses",
    x = "Saves", fill = "Game Result"
  ) +
  theme_minimal()
```

Just asked if it could analyze ground balls' correlation to wins instead, gave this.

```{r}
lacrosse_games2025 %>%
  group_by(result) %>%
  summarize(
    avg_ground_balls = mean(gb, na.rm = TRUE),
    median_ground_balls = median(gb, na.rm = TRUE),
    count = n()
  )
```

Asked if it could analyze turnovers' correlation to wins, gave this.

```{r}

lacrosse_games2025 %>%
  group_by(result) %>%
  summarize(
    avg_turnovers = mean(to, na.rm = TRUE),
    median_turnovers = median(to, na.rm = TRUE),
    count = n()
  )
```
```{r}
ggplot(lacrosse_games2025, aes(x = to, fill = result)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 25) +
  labs(
    title = "Turnovers in Wins vs. Losses",
    x = "Turnovers", fill = "Game Result"
  ) +
  theme_minimal()
```
Asked if it could analyze shots' correlation to wins instead, gave this.
```{r}
lacrosse_games2025 %>%
  group_by(result) %>%
  summarize(
    avg_shots = mean(shots, na.rm = TRUE),
    median_shots = median(shots, na.rm = TRUE),
    count = n()
  )
```








